\documentclass[12pt,a4paper]{article}

% Packages
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{physics}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage[margin=2.5cm]{geometry}
\usepackage{enumerate}
\usepackage{float}
\usepackage{booktabs}
\usepackage{natbib}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\kB}{k_{\mathrm{B}}}
\newcommand{\Sosc}{S_{\mathrm{osc}}}
\newcommand{\Scat}{S_{\mathrm{cat}}}
\newcommand{\Spart}{S_{\mathrm{part}}}
\newcommand{\Stotal}{S_{\mathrm{total}}}

\title{On the Thermodynamic Consequence of the Equivalence in Oscillatory, Categorical, and Partitioning Representation: Mechanistic Synthesis of Sequential Operations for the Resolution of Irreversible Operations}
\author{
Kundai Farai Sachikonye\\
\texttt{kundai.sachikonye@wzw.tum.de}
}

\begin{document}

\maketitle

\begin{abstract}
We present a unified thermodynamic framework demonstrating that oscillatory dynamics, categorical structure, and partition operations yield identical entropy formulations when derived from independent first principles. Beginning with three separate derivations—entropy from bounded oscillatory systems, entropy from categorical state spaces, and entropy from partition branching structures—we prove that all three converge to the same formula $S = \kB M \ln n$, where $M$ represents the dimensional depth and $n$ the branching factor. This convergence establishes a fundamental equivalence: oscillation, category, and partition are not merely related phenomena but identical structures viewed from different perspectives. We then introduce \emph{partition lag}—the irreducible temporal gap between the act of partitioning and the partitioned result—and demonstrate that this lag generates entropy through undetermined residue. The key result is that partition operations are thermodynamically irreversible: composition cannot recover the entropy lost to partition boundaries. We apply this framework to analyse physical systems including finite geometric partitioning of aggregate properties, infinite subdivision of bounded continuous intervals, continuous-to-discrete temporal decomposition, identity persistence under sequential component exchange, partition-free traversal of continuous intervals, and non-partitionable accumulation of resolved alternatives. The thermodynamic analysis reveals that certain classical puzzles in philosophy and physics dissolve when recognised as consequences of partition-induced entropy production. Notably, the framework provides partition-theoretic derivations of null geodesics (maximum speed as partition-free traversal), temporal duration (proper time as accumulated partition entropy), and the cosmological dark-to-ordinary matter ratio (non-partitionable resolved alternatives). We further establish that non-actualisations possess intrinsic geometric structure organised by categorical distance, with close non-actualisations forming paired reference structures (ordinary matter) and distant non-actualisations remaining unpaired (dark matter). Finally, we prove that actualisation is logically prior to non-actualisation—every negation presupposes what it negates—resolving the question of why there is something rather than nothing.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

The relationship between microscopic dynamics and macroscopic thermodynamics has been a central problem in physics since Boltzmann's statistical interpretation of entropy. We present a framework that unifies three apparently distinct approaches to this relationship: oscillatory mechanics, categorical enumeration, and partition theory. Our central result is that these three approaches, when developed from independent axioms, yield identical entropy formulations—demonstrating not merely an analogy but a fundamental equivalence.

The paper proceeds in three parts. In Part I, we derive entropy independently from oscillatory, categorical, and partition perspectives, then prove their mathematical equivalence. In Part II, we introduce partition lag and demonstrate that partition operations generate irreversible entropy through undetermined residue. In Part III, we apply the framework to physical systems, revealing thermodynamic resolutions to problems in mechanics and ontology.

Throughout, we employ standard thermodynamic notation with Boltzmann's constant $\kB = 1.380649 \times 10^{-23}$ J/K explicit, emphasising that our results concern physical entropy rather than abstract information measures.

%============================================================
% PART I: INDEPENDENT ENTROPY DERIVATIONS
%============================================================

\part{Entropy Unification}
\label{part:unification}

\input{sections/oscillatory-mechanics}
\input{sections/categorical-mechanics}
\input{sections/partitioning-mechanisms}
\input{sections/entropy-unification}

%============================================================
% PART II: PARTITION LAG AND IRREVERSIBILITY
%============================================================

\part{Partition Lag and Irreversibility}
\label{part:partition_lag}

\input{sections/partition-lag}

%============================================================
% PART III: PHYSICAL APPLICATIONS
%============================================================

\part{Physical Applications}
\label{part:applications}

\input{sections/heap-paradox}
\input{sections/zenos-paradox}
\input{sections/ship-of-theseus}
\input{sections/null-geodesics}
\input{sections/recursive-compounding}
\input{sections/geometry-of-non-actualisation}
\input{sections/priority-of-existence}
\input{sections/hydrogen-from-negation}

\section{Conclusion}
\label{sec:conclusion}

We have established three independent derivations of entropy—from oscillatory mechanics, categorical structure, and partition theory—and proved their mathematical equivalence. The unified entropy formula $S = \kB M \ln n$ emerges identically from all three perspectives, demonstrating that oscillation, category, and partition are not analogous but identical.

The partition lag mechanism reveals why composition cannot reverse partition: each partition operation generates undetermined residue that increases entropy by $\Delta S > 0$. This irreversibility is not a limitation of particular physical systems but a consequence of the fundamental structure of categorical operations.

The physical applications demonstrate that systems traditionally analysed through composition—asking how parts combine to form wholes—are more naturally understood through partition—asking how wholes decompose into parts with entropy loss. The thermodynamic framework provides quantitative predictions for the entropy cost of partition and explains why certain properties of wholes cannot be recovered from parts.

Two results concerning light and dark matter emerge from the partition framework. First, partition-free traversal—motion without creating categorical distinctions—generates zero boundary entropy and therefore zero proper time, providing a thermodynamic derivation of null geodesics and explaining why the speed of light is maximum. Second, each actualisation resolves infinitely many non-actualisations into ``did not happen,'' and these non-actualisations cannot be partitioned (they lack categorical structure), hence cannot interact with partition-free entities (light), hence are ``dark'' while still contributing gravitationally.

The geometric structure of non-actualisation space provides the quantitative foundation for the dark matter ratio. Non-actualisations are organised in shells of increasing categorical distance from actualisations. Close non-actualisations pair with nearby actualisations to form mutual reference structures—this pairing IS the relational structure of ordinary matter. Distant non-actualisations lack pairing partners and remain as unstructured, non-partitionable mass. The $\sim 5:1$ ratio of unpaired to paired non-actualisations emerges from the exponential growth of shells with distance, determined by the branching factor of categorical space.

Finally, the logical priority of actualisation resolves the question of why there is something rather than nothing. Every negation presupposes what it negates: ``not-$X$'' requires $X$ to exist as its referent. Non-actualisations depend ontologically on actualisations, not vice versa. Pure nothing—the absence of all actualisation—is self-contradictory, because ``nothing exists'' is itself a determination that presupposes the meaningfulness of ``something.'' Existence is logically necessary; non-existence is derivative and dependent.

The partition framework thus unifies thermodynamics, relativity, cosmology, and ontology under a single principle: the categorical structure of distinction-making determines what can exist (partitionable), what can move at maximum speed (partition-free), what can interact (at least one participant partitions), what remains invisible yet gravitating (non-partitionable), and why existence precedes non-existence (negation presupposes affirmation).

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}

