\documentclass[11pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{mathrsfs}
\usepackage{xcolor}
\usepackage{siunitx}
\usepackage{float}          % Better figure placement
\usepackage{caption}        % Better captions
\usepackage{subcaption}     % Subfigures (if needed)

% Theorem environments
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}{Definition}

\newcommand{\Sk}{S_k}           % Kinetic entropy
\newcommand{\St}{S_t}           % Topological entropy
\newcommand{\Se}{S_e}           % Evolutionary entropy
\newcommand{\Sspace}{\mathcal{S}}  % S-space
\newcommand{\kB}{k_\text{B}}    % Boltzmann constant


% Bold vectors
\renewcommand{\vec}[1]{\mathbf{#1}}

% Expectation value
\newcommand{\expect}[1]{\langle #1 \rangle}

% Differential operator
\newcommand{\diff}{\mathrm{d}}

\title{On The Consequences of Categorical Completion on Ensemble Membrane Transporters: Mechanistic Synthesis of Transport Mechanisms as Ensemble Membrane Maxwell Demons}

\author{Kundai Sachikonye}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Membrane transporters have been theoretically identified as molecular Maxwell demons that maintain concentration gradients through information-driven selection. We establish the mechanistic basis for this behavior through phase-locking of terahertz-frequency vibrational modes and validate it via trans-Planckian observation with zero quantum backaction. We map transporter conformational states to S-entropy coordinates in categorical space, demonstrate that substrate selection emerges from frequency matching in the range \SI{3.2e13}{}-\SI{4.5e13}{\hertz}, and prove that ATP hydrolysis modulates binding site frequencies to scan for substrates. Computational validation on five test substrates shows selectivity factors of \num{9.1e9} for individual transporters. We extend the framework to ensemble behavior, modeling \num{5000} transporters as a single collective demon that exhibits emergent properties: enhanced throughput (\SI{42500}{molecules\per\second}), statistical frequency coverage, and sharpened selectivity through ensemble averaging. Trans-Planckian observations at femtosecond resolution confirm zero momentum transfer (\SI{0.00}{\kilogram\meter\per\second}) across \num{300} measurements, validating categorical measurement without quantum backaction. The ensemble demon successfully discriminates between substrates in multi-substrate competition (collective selectivity \num{1e10}), with weak substrates showing \SI{72}{\percent} efficiency versus \SI{100}{\percent} for strong substrates. These results establish membrane transporters as phase-locked categorical Maxwell demons operating through dual physical-categorical coordinate systems, with ensemble behavior emerging from collective dynamics in S-entropy space.
\end{abstract}

\section{Introduction}

Membrane transporters maintain non-equilibrium concentration gradients that are essential for cellular function. ATP-binding cassette (ABC) transporters, representing a major family with \num{49} human members, actively export or import substrates across membranes using ATP hydrolysis energy~\cite{dean2001human,locher2016mechanistic}. Flatt et al.~\cite{flatt2023abc} recently demonstrated that ABC transporters constitute physical realisations of Maxwell demons, theoretical devices that use information to sort molecules and maintain gradients without direct mechanical work~\cite{maxwell1871theory,bennett1982thermodynamics}.

While the information-theoretic framework establishes transporters as Maxwell demons, the mechanistic basis remains unclear. How do transporters "measure" substrates? What physical process constitutes "information processing"? How does ATP provide the energy for information erasure required by Landauer's principle~\cite{landauer1961irreversibility}?

We address these questions through three advances:

\textbf{(1) Mechanistic explanation:} We show substrate selection emerges from phase-locking between binding site vibrational frequencies (\SI{e13}{\hertz} range) and substrate molecular vibrations, not geometric lock-and-key recognition alone.

\textbf{(2) Categorical coordinates:} We map conformational states to S-entropy coordinates in categorical space orthogonal to physical space, enabling information extraction without physical disturbance.

\textbf{(3) Ensemble demon:} We model all transporters of one type as a collective demon exhibiting emergent properties: enhanced throughput, statistical coverage, and collective selectivity.

Computational validation demonstrates selectivity factors of \num{9e9}, transport rates of \SI{42500}{molecules\per\second} for \num{5000}-transporter ensembles, and zero quantum backaction (\SI{0.00}{\kilogram\meter\per\second}) in femtosecond-resolution observations.

\section{Membrane Transporter Maxwell Demon}
\label{sec:demon}
\input{sections/membrane-transporter-maxwell-demon}

\section{Categorical Coordinate Space}
\label{sec:categorical}
\input{sections/categorical-coordinate-space}

\section{Phase-Locked Substrate Selection}
\label{sec:phase-lock}
\input{sections/phase-locked-substrate-selection}

\section{Zero-Backaction Observation}
\label{sec:observation}
\input{sections/zero-backaction-observation}

\section{Ensemble Transporter Demon}
\label{sec:ensemble}
\input{sections/ensemble-transporter-demon}

\section{Experimental Validation}
\label{sec:validation}
\input{sections/experimental-validation}

\section{Conclusions}

We have established membrane transporters as phase-locked categorical Maxwell demons operating through dual physical-categorical coordinate systems:

\textbf{Individual transporter:} Conformational states map to S-entropy coordinates (S-space distance traveled: \num{14.73} over 5 ATP cycles). Substrate selection occurs through phase-locking in the \SI{3.2e13}{}-\SI{4.5e13}{\hertz} range (threshold: \SI{e12}{\hertz}). ATP modulates binding site frequency over \SI{1.3e13}{\hertz} range, enabling frequency scanning. Five test substrates show selectivity factor \num{9.1e9}, with Verapamil transported (phase-lock: \num{0.910}) and four substrates rejected (average phase-lock: \num{0.154}).

\textbf{Ensemble demon:} \num{5000} transporters modeled as collective demon exhibit enhanced throughput (\SI{42500}{molecules\per\second}, \num{100}-fold above statistical expectation), continuous frequency coverage through distributed ATP cycles, and collective selectivity (\num{1e10}). Multi-substrate competition shows efficiency discrimination: weak substrates \SI{72}{\percent}, strong substrates \SI{100}{\percent}.

\textbf{Trans-Planckian observation:} Femtosecond-resolution (\SI{e-15}{\second}) observations across \num{300} measurements yield exactly zero momentum transfer (\SI{0.00}{\kilogram\meter\per\second}), confirming categorical measurement without quantum backaction. Backaction-to-Heisenberg ratio: \num{0.00}; backaction-to-thermal ratio: \num{0.00}.

These results validate the information-theoretic Maxwell demon framework through mechanistic phase-locking dynamics and demonstrate that ensemble behavior emerges from collective operation in categorical space. The framework applies to all ATP-driven membrane transporters and explains substrate promiscuity, drug resistance mechanisms, and membrane domain effects through frequency-space dynamics.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
