\documentclass[12pt,a4paper]{article}

% Packages
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{physics}
\usepackage{siunitx}
\usepackage{algorithm}
\usepackage{algpseudocode}

\geometry{margin=2.5cm}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}

% Title
\title{Partition Coordinate Geometry in Bounded Oscillatory Systems}
\author{}
\date{}

\begin{document}

\maketitle

\begin{abstract}
We develop a mathematical framework for describing categorical states in bounded phase spaces using \emph{partition coordinates}---a four-parameter addressing system arising from nested boundary constraints. We prove that the geometry of bounded partitioning imposes strict constraints on coordinate values: the depth parameter $n \geq 1$, the complexity parameter $l \in \{0, 1, \ldots, n-1\}$, the orientation parameter $m \in \{-l, \ldots, +l\}$, and a binary chirality parameter $s \in \{-\frac{1}{2}, +\frac{1}{2}\}$. From these constraints alone, we derive a fundamental capacity theorem: the maximum number of distinct states at partition depth $n$ is exactly $2n^2$. We show that energy minimisation in partition space produces a specific filling order, and that transitions between coordinates follow selection rules determined by boundary continuity. We extend the theory to multi-body systems where both boundaries and centers carry chirality, deriving hyperfine splitting from chirality-chirality coupling. For the simplest partition configuration, this predicts a hyperfine transition at 1420 MHz (21 cm wavelength). Hardware-based virtual instruments are constructed to measure partition coordinates, and a Universal Virtual Instrument Finder algorithm systematically constructs optimal measurement configurations from arbitrary hardware. We extend the framework to multi-atom systems, defining \emph{partition signatures} that uniquely characterise compounds. This enables mixture identification, compound feasibility prediction, and de novo molecular design---all from partition geometry. The resulting framework exhibits structural similarities to several known physical systems, which we note in concluding remarks.
\end{abstract}

\tableofcontents
\newpage

%==============================================================================
\part{Mathematical Foundations}
\label{part:foundations}
%==============================================================================

\input{sections/partition-coordinates}
\input{sections/geometric-constraints}
\input{sections/energy-ordering}

%==============================================================================
\part{Measurement Theory}
\label{part:measurement}
%==============================================================================

\input{sections/hardware-based-partitioning-coordinates}
\input{sections/spectral-transitions}
\input{sections/property-trends}

%==============================================================================
\part{Experimental Validation}
\label{part:experimental}
%==============================================================================

\input{sections/virtual-instrument-measurements}
\input{sections/coordinate-uniqueness}

%==============================================================================
\part{Extended Theory}
\label{part:extended}
%==============================================================================

\input{sections/nuclear-magnetic-resonance}
\input{sections/mass-spectrometry-ensemble}
\input{sections/categorical-state-synthesizer}
\input{sections/instrument-orchestration}
\input{sections/universal-instrument-algorithm}
\input{sections/compound-identification-design}

%==============================================================================
\part{Discussion}
\label{part:discussion}
%==============================================================================

\section{Summary of Results}
\label{sec:summary}

We have established the following results from first principles:

\begin{enumerate}
    \item \textbf{Partition coordinates} $(n, l, m, s)$ provide a complete addressing system for categorical states in bounded phase space.

    \item \textbf{Geometric constraints} arise from boundary nesting: $l < n$, $|m| \leq l$, $s = \pm\frac{1}{2}$.

    \item \textbf{Capacity formula}: The number of distinct states at depth $n$ is exactly $2n^2$.

    \item \textbf{Energy ordering} follows the $(n + \alpha l)$ rule, producing a specific filling sequence.

    \item \textbf{Transition rules} follow from boundary continuity: $\Delta l = \pm 1$, $\Delta m \in \{0, \pm 1\}$, $\Delta s = 0$.

    \item \textbf{Coordinate uniqueness}: No two categorical states can share identical coordinates.

    \item \textbf{Property trends}: Measurable quantities show systematic variation across partition space.

    \item \textbf{Hyperfine structure}: Center chirality $s_c$ couples to boundary chirality $s$, producing energy splitting of $\Delta E_{\text{hf}} = 5.87 \times 10^{-6}$ eV for the ground state of $Z = 1$.

    \item \textbf{Element signatures}: Each partition count $Z$ has a unique coordinate configuration that determines all observable properties.

    \item \textbf{Instrument equivalence}: Partition coordinates can be measured by four independent instrument categories---exotic partition instruments, standard chemistry instruments (mass spectrometry, XPS, NMR, ESR), virtual spectrometers (UV-Vis, IR, Raman), and computational methods---all yielding identical results.

    \item \textbf{Categorical instrument orchestration}: The instrument ensemble constitutes a Poincar\'{e} machine---solutions are trajectories through instrument space that achieve recurrence (all projections agree). Element identification has Poincar\'{e} complexity $\Pi(Z) \in [2, 5]$ depending on configuration complexity.

    \item \textbf{Universal Virtual Instrument Algorithm}: A systematic procedure (Algorithm~\ref{alg:uvif}) constructs optimal virtual instruments from arbitrary hardware. Given available oscillators, target coordinates, and precision requirements, the algorithm outputs an optimal instrument configuration, measurement protocol, and coordinate extraction procedure with computational complexity $\mathcal{O}(N \cdot M \cdot |\Omega| + 2^N \cdot M)$.

    \item \textbf{Partition signatures}: Every compound has a unique partition signature $\Sigma(M)$---the multiset of all constituent electron coordinates. Two compounds are identical if and only if their signatures match (Theorem~\ref{thm:signature_uniqueness}).

    \item \textbf{Mixture identification}: Unknown mixtures can be decomposed into constituent compounds by analysing their combined partition signature (Algorithm~\ref{alg:mixture_identification}).

    \item \textbf{Compound feasibility}: The feasibility of proposed compounds can be predicted from partition coordinate constraints---whether valid bonding configurations exist that minimise energy (Algorithm~\ref{alg:feasibility}).

    \item \textbf{De novo design}: New compounds with target properties can be designed by searching partition coordinate space for signatures that produce desired characteristics (Algorithm~\ref{alg:denovo_design}).
\end{enumerate}

\section{Structural Correspondences}
\label{sec:correspondences}

The mathematical structure derived here exhibits notable correspondences with known physical systems:

\begin{itemize}
    \item The partition coordinates $(n, l, m, s)$ have the same constraint structure as the quantum numbers $(n, l, m_l, m_s)$ of atomic physics.

    \item The capacity formula $2n^2$ matches the electron shell capacity in atoms.

    \item The energy ordering reproduces the aufbau filling principle.

    \item The transition selection rules match atomic spectral selection rules.

    \item The coordinate uniqueness principle has the same form as the Pauli exclusion principle.

    \item The property trends across partition space mirror periodic trends in chemistry.

    \item The hyperfine splitting from chirality coupling matches the hydrogen 21 cm line (1420.405 MHz), used extensively in radio astronomy.

    \item The center chirality measurement corresponds to nuclear magnetic resonance (NMR) spectroscopy.

    \item Standard analytical instruments (mass spectrometry, XPS, ESR, UV-Vis) already measure partition coordinates---they extract $(n, l, m, s)$ values without knowing they are doing so.
\end{itemize}

These correspondences suggest that atomic structure may be a physical instantiation of partition coordinate geometry. If this interpretation is correct, then:

\begin{enumerate}
    \item The periodic table is a geometric necessity rather than an empirical accident.

    \item Chemical elements are defined by their partition coordinate signatures.

    \item Spectroscopy measures transitions between partition coordinates.

    \item Chemical properties emerge from the geometry of bounded phase space.

    \item Standard chemistry instrumentation has been measuring partition geometry all along---the partition framework provides a unified interpretation of diverse measurements.

    \item Analytical chemistry is a form of Poincar\'{e} Computing---element identification is trajectory completion through instrument space, with solution recognised when all projections achieve recurrence.

    \item Molecular compounds are characterised by their partition signatures---the multiset of all electron coordinates. This provides a complete molecular fingerprint.

    \item Drug design, materials discovery, and chemical synthesis can be reformulated as optimisation problems in partition coordinate space.
\end{enumerate}

We leave detailed investigation of these correspondences to future work.

\section{Conclusion}
\label{sec:conclusion}

We have developed a complete mathematical theory of partition coordinates in bounded oscillatory systems. The theory is self-contained, requiring only the axioms of categorical partitioning and the constraint of bounded phase space. All results---including the $2n^2$ capacity formula, energy ordering, transition rules, and uniqueness principle---follow from geometry alone.

The structural similarity between this framework and atomic physics is striking but not presupposed. We have not assumed any knowledge of quantum mechanics or chemistry in our derivations. The correspondences emerge as consequences of the mathematics.

Whether these correspondences indicate a deep connection between categorical partitioning and the fundamental structure of matter remains an open question. The framework presented here provides the mathematical tools to investigate this possibility.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

